---
title: "Kathryn"
author: "Kathryn Yalch"
date: "2024-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
par(pin = c(12, 8))
```

```{r}
# Set working directory
setwd("C:/Users/kyalc/Downloads/pitcheranalysis/pitcheranalysis")

# 2021 Season
gamescore21 <- read.csv("data/gamescore_by_gameid_ab/gameid_ab_score_21.csv")
pitches21 <- read.csv("data/pitches_21.csv")
pitchwithscore21 <- left_join(pitches21, gamescore21, by = c("gameid", "ab")) # joining score to pitches21
write_csv(pitchwithscore21, "data/pitchesWithScore21.csv")

# 2022 Season
gamescore22 <- read.csv("data/gamescore_by_gameid_ab/gameid_ab_score_22.csv")
pitches22 <- read.csv("data/pitches_22.csv")
pitchwithscore22 <- left_join(pitches22, gamescore22, by = c("gameid", "ab")) # joining score to pitches22
write_csv(pitchwithscore22, "data/pitchesWithScore22.csv")

# 2023 Season
gamescore23 <- read.csv("data/gamescore_by_gameid_ab/gameid_ab_score_23.csv")
pitches23 <- read.csv("data/pitches_23.csv")
pitchwithscore23 <- left_join(pitches23, gamescore23, by = c("gameid", "ab")) # joining score to pitches23
write_csv(pitchwithscore23, "data/pitchesWithScore23.csv")

# Reading back the merged files
pitchwithscore21 <- read_csv("data/pitchesWithScore21.csv", show_col_types = FALSE)
pitchwithscore22 <- read_csv("data/pitchesWithScore22.csv", show_col_types = FALSE)
pitchwithscore23 <- read_csv("data/pitchesWithScore23.csv", show_col_types = FALSE)


```

## Closer EDA

#### Build Closer Datasets Per Year

```{r}
# Closers | 2021 season 
pitchwithscore21$closeness <- abs(pitchwithscore21$homscore - pitchwithscore21$visscore)
#filtering by closeness and whether the pitcher is closing
datawscore21<- filter(pitchwithscore21, closeness<=3, inning>=8)
#examining espn's top closers
espnbest21 <- c("Melancon, Mark", "Hendriks, Liam", "Jansen, Kenley", "Smith, Will", "Iglesias, Raisel", "Hader, Josh", "Diaz, Edwin", "McGee, Jake", "Chapman, Aroldis", "Reyes, Alex", "Pressly, Ryan", "Clase, Emmanuel", "Barnes, Matt", "Romano, Jordan", "Kimbrel, Craig", "Trivino, Lou", "Hand, Brad", "Bard, Daniel", "Soto, Gregory","Colome, Alex", "Barlow, Scott", "Kennedy, Ian", "Floro, Dylan", "Steckenrider, Drew", "Rodriguez, Richard", "Castillo, Diego", "Neris, Hector", "Sulser, Cole", "Hembree, Heath", "Soria, Joakim")
#above from https://www.espn.com/mlb/stats/closers/_/year/2021
espnsbestclosers21<- filter(pitchwithscore21, closeness<=3, inning>=8, pitcher %in% espnbest21)
```

```{r}
# Closers | 2022 season
pitchwithscore22$closeness <- abs(pitchwithscore22$homscore - pitchwithscore22$visscore)
#filtering by closeness and whether the pitcher is closing
datawscore22<- filter(pitchwithscore22, closeness<=3, inning>=8)
#examining espn's top closers
espnbest22 <- c("Clase, Emmanuel", "Jansen, Kenley", "Hendriks, Liam", "Romano, Jordan", "Bard, Daniel", "Pressly, Ryan", "Diaz, Edwin", "Soto, Gregory", "Hader, Josh", "Rogers, Taylor", "Doval, Camilo", "Barlow, Scott", "Kimbrel, Craig", "Sewald, Paul", "Holmes, Clay", "Scott, Tanner", "Helsley, Ryan", "Bednar, David", "Lopez, Jorge", "Melancon, Mark", "Iglesias, Raisel", "Robertson, David", "Barlow, Joe", "Knebel, Corey", "Rainey, Tanner", "Jimenez, Dany", "Diaz, Alexis", "Pagan, Emilio", "Schreiber, John")
espnsbestclosers22<- filter(pitchwithscore22, closeness<=3, inning>=8, pitcher %in% espnbest22)
```

```{r}
# Closers | 2023 season
pitchwithscore23$closeness <- abs(pitchwithscore23$homscore - pitchwithscore23$visscore)
#filtering by closeness and whether the pitcher is closing
datawscore23<- filter(pitchwithscore23, closeness<=3, inning>=8)
#examining espn's top closers
espnbest23 <- c("Clase, Emmanuel", "Bednar, David", "Doval, Camilo", "Diaz, Alexis", "Romano, Jordan", "Williams, Devin", "Bautista, Felix", "Hader, Josh", "Iglesias, Raisel", "Pressly, Ryan", "Estevez, Carlos", "Jansen, Kenley", "Finnegan, Kyle", "Duran, Jhoan", "Lange, Alex", "Fairbanks, Pete", "Holmes, Clay", "Phillips, Evan", "Kimbrel, Craig", "Alzolay, Adbert", "Smith, Will", "May, Trevor", "Sewald, Paul", "Puk, A.J.", "Robertson, David", "Helsley, Ryan", "Johnson, Pierce", "Barlow, Scott", "Sewald, Paul", "Graveman, Kendall")
espnsbestclosers23<- filter(pitchwithscore23, closeness<=3, inning>=8, pitcher %in% espnbest23)
```

#### Combined Dataset

```{r}
all_closer_data<- rbind(espnsbestclosers21, espnsbestclosers22, espnsbestclosers23) %>% select(-X)
```

### Closer Pitch Type Frequency

```{r}
# ALL CLOSER DATA: Get each pitch type, plot frequencies, boxplot for pitching stats
all_closer_pitchtypes <- unique(all_closer_data$pitchname)
all_closer_pitchdesc <- unique(all_closer_data$pitchname_desc)

ggplot(data=all_closer_data, mapping = aes(x = pitchname, fill = pitchname))  +
  geom_bar() + xlab("Pitch Type") + ylab("Frequency") + ggtitle("Pitch Type Frequency for Closers from 2021-2023")
# Four-seam FB
all_closer_ff <- filter(all_closer_data, pitchname=="FF")
boxplot(scale(select(all_closer_ff, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight), center = FALSE), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Fastball Pitching Stats Boxplots")
# Sinker
all_closer_si <- filter(all_closer_data, pitchname=="SI")
boxplot(scale(select(all_closer_si, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Sinker Pitching Stats Boxplots")
# Slider
all_closer_sl <- filter(all_closer_data, pitchname=="SL")
boxplot(scale(select(all_closer_sl, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Slider Pitching Stats Boxplots")
# Curveball
all_closer_cu <- filter(all_closer_data, pitchname=="CU")
boxplot(scale(select(all_closer_cu, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Curveball Pitching Stats Boxplots")
# Changeup
all_closer_ch <- filter(all_closer_data, pitchname=="CH")
boxplot(scale(select(all_closer_ch, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Changeup Pitching Stats Boxplots")
# Cutter
all_closer_fc <- filter(all_closer_data, pitchname=="FC")
boxplot(scale(select(all_closer_fc, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Cutter Pitching Stats Boxplots")
# Splitter
all_closer_fs <- filter(all_closer_data, pitchname=="FS")
boxplot(scale(select(all_closer_fs, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Splitter Pitching Stats Boxplots")
# Knuckle Curve
all_closer_kc <- filter(all_closer_data, pitchname=="KC")
boxplot(scale(select(all_closer_kc, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Knuckle Curve Pitching Stats Boxplots")
# Sweeper
all_closer_st <- filter(all_closer_data, pitchname=="ST")
boxplot(scale(select(all_closer_st, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Sweeper Pitching Stats Boxplots")
```

```{r}
all_closer_pitchres <- unique(all_closer_data$pitchresult)
all_closer_pitchres_desc <- na.omit(unique(all_closer_data$pitchresult_desc))

data.frame(all_closer_pitchres,all_closer_pitchres_desc)
# Strike/Out: C, F, S, T, W, L, O, M, X
strike_out <- c("C","F","S","T","W","L","O","M","X")
# Ball:
```

```{r}
out_closer_pitch <- all_closer_data %>% filter(pitchresult %in% strike_out)
ggplot(data=out_closer_pitch, mapping = aes(x = pitchname, fill = pitchname))  + geom_bar() + xlab("Pitch Type") + ylab("Frequency") + ggtitle("Closers Pitch Type Frequency Resulting in Strikeout/Outs")
# Four-seam FB
all_closer_ff <- filter(all_closer_data, pitchname=="FF", pitchresult %in% strike_out)
boxplot(scale(select(all_closer_ff, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Fastball Pitching Stats Boxplots")
# Sinker
all_closer_si <- filter(all_closer_data, pitchname=="SI", pitchresult %in% strike_out)
boxplot(scale(select(all_closer_si, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Sinker Pitching Stats Boxplots")
# Slider
all_closer_sl <- filter(all_closer_data, pitchname=="SL", pitchresult %in% strike_out)
boxplot(scale(select(all_closer_sl, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Slider Pitching Stats Boxplots")
# Curveball
all_closer_cu <- filter(all_closer_data, pitchname=="CU", pitchresult %in% strike_out)
boxplot(scale(select(all_closer_cu, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Curveball Pitching Stats Boxplots")
# Changeup
all_closer_ch <- filter(all_closer_data, pitchname=="CH", pitchresult %in% strike_out)
boxplot(scale(select(all_closer_ch, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Changeup Pitching Stats Boxplots")
# Cutter
all_closer_fc <- filter(all_closer_data, pitchname=="FC", pitchresult %in% strike_out)
boxplot(scale(select(all_closer_fc, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Cutter Pitching Stats Boxplots")
# Splitter
all_closer_fs <- filter(all_closer_data, pitchname=="FS", pitchresult %in% strike_out)
boxplot(scale(select(all_closer_fs, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Splitter Pitching Stats Boxplots")
# Knuckle Curve
all_closer_kc <- filter(all_closer_data, pitchname=="KC", pitchresult %in% strike_out)
boxplot(scale(select(all_closer_kc, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Knuckle Curve Pitching Stats Boxplots")
# Sweeper
all_closer_st <- filter(all_closer_data, pitchname=="ST", pitchresult %in% strike_out)
boxplot(scale(select(all_closer_st, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Sweeper Pitching Stats Boxplots")
```

### Individual Closer Datasets

```{r}
# vector of all closing pitchers (not used right now)
every_closer <- unique(c(espnbest21, espnbest22, espnbest23)) 

pitcherName <- paste0(all_closer_data$pitcher, " table")

#closer_tables is a list of tables
closer_tables <- split(all_closer_data,pitcherName)

for (i in closer_tables) {
  pitchtypes <- unique(i$pitchname)
  pitchdesc <- unique(i$pitchname_desc)
  title_name <- paste((i[1,10]), "Pitch Type Frequency", sep = " ")
  print(ggplot(data=i, mapping = aes(x = pitchname))  +
  geom_bar() + xlab("Pitch Type") + ylab("Frequency") + ggtitle(title_name))
}

```

#### Individual Closer Datasets

```{r}
# Way 1: Reliever Pitch Type Medians per row
pitcherName <- paste0(all_closer_data$pitcher, " table")
closer_tables <- split(all_closer_data, pitcherName)

# Create an empty data frame to store averages
median_by_closer <- data.frame(name=character(), pitch=character(), median_spin=numeric(), median_velo=numeric(),
                              median_hbreak=numeric(), median_vbreak=numeric(), median_plside=numeric(),
                              median_plheight=numeric(), stringsAsFactors=FALSE)

# Get Average Stats for Each Pitcher's Pitch Types

# Loop through each reliever
for (i in closer_tables) {
  pitchtypes <- unique(i$pitchname)

  # Split by pitch type
  table_by_pitch <- split(i, i$pitchname)

  # Loop through each pitch type and calculate averages
  for (j in table_by_pitch) {
    median_spin <- median(j$spinrate, na.rm=TRUE)
    median_velo <- median(j$relspeed, na.rm=TRUE)
    median_hbreak <- median(j$horzbreak, na.rm=TRUE)
    median_vbreak <- median(j$inducedvertbreak, na.rm=TRUE)
    median_plside <- median(j$platelocside, na.rm=TRUE)
    median_plheight <- median(j$platelocheight, na.rm=TRUE)

    # Append the data to median_by_reliever
    median_by_closer <- rbind(median_by_closer,
                             data.frame(name=i$pitcher[1], pitch=j$pitchname[1], median_spin=median_spin,
                                        median_velo=median_velo, median_hbreak=median_hbreak, median_vbreak=median_vbreak,
                                        median_plside=median_plside, median_plheight=median_plheight))
  }
}

# plot
ggplot(median_by_closer, aes(x=pitch, y=median_vbreak)) +
  geom_boxplot() +
  ggtitle("Closer Boxplots of Vertical Break by Pitch Type")
```

#### Closer Table

```{r}
find_outlier <- function(x) {
  return(x < quantile(x, .25) - 1.5*IQR(x) | x > quantile(x, .75) + 1.5*IQR(x))
}
median_by_pitcher <- median_by_pitcher %>%
        group_by(pitch) %>%
        mutate(outlier = ifelse(find_outlier(median_spin), name, NA))
ggplot(median_by_pitcher, aes(x=pitch, y=median_spin)) +
  geom_boxplot() +
  ggtitle("Boxplots of Spin Rates by Pitch")+
  geom_text(aes(label=outlier), na.rm=TRUE, hjust=-.5)
median_by_pitcher <- median_by_pitcher %>%
        group_by(pitch) %>%
        mutate(outlier = ifelse(find_outlier(median_velo), name, NA))
ggplot(median_by_pitcher, aes(x=pitch, y=median_velo)) +
  geom_boxplot() +
  ggtitle("Boxplots of Velocity by Pitch")+
  geom_text(aes(label=outlier), na.rm=TRUE, hjust=-.5)
median_by_pitcher <- median_by_pitcher %>%
        group_by(pitch) %>%
        mutate(outlier = ifelse(find_outlier(median_hbreak), name, NA))
ggplot(median_by_pitcher, aes(x=pitch, y=median_hbreak)) +
  geom_boxplot() +
  ggtitle("Boxplots of Horizontal Break by Pitch")+
  geom_text(aes(label=outlier), na.rm=TRUE, hjust=-.5)
median_by_pitcher <- median_by_pitcher %>%
        group_by(pitch) %>%
        mutate(outlier = ifelse(find_outlier(median_vbreak), name, NA))
ggplot(median_by_pitcher, aes(x=pitch, y=median_vbreak)) +
  geom_boxplot() +
  ggtitle("Boxplots of Vertical Break by Pitch")+
  geom_text(aes(label=outlier), na.rm=TRUE, hjust=-.5)
median_by_pitcher <- median_by_pitcher %>%
        group_by(pitch) %>%
        mutate(outlier = ifelse(find_outlier(median_plside), name, NA))
ggplot(median_by_pitcher, aes(x=pitch, y=median_plside)) +
  geom_boxplot() +
  ggtitle("Boxplots of Vertical Location by Pitch")+
  geom_text(aes(label=outlier), na.rm=TRUE, hjust=-.5)
median_by_pitcher <- median_by_pitcher %>%
        group_by(pitch) %>%
        mutate(outlier = ifelse(find_outlier(median_plheight), name, NA))
ggplot(median_by_pitcher, aes(x=pitch, y=median_plheight)) +
  geom_boxplot() +
  ggtitle("Boxplots of Height by Pitch")+
  geom_text(aes(label=outlier), na.rm=TRUE, hjust=-.5)
```

#### Closer Pitch Location & Result by Pitch Type

```{r}
#location of pitches groups by pitch type
ggplot(datawscore21, aes(x=platelocside, y= platelocheight)) + geom_point(alpha = 0.1)+ 
  facet_grid(pitchname~.)

#examining if the location of pitches is dependent on handedness of pitcher
ggplot(datawscore21, aes(x=platelocside, y= platelocheight)) + geom_point(alpha = 0.1)+ 
  facet_grid(pitchname~pitcherthrows)

#are any ambidextrous? no
sum(datawscore21$pitcherthrows=="R")
sum(datawscore21$pitcherthrows=="L")
sum(datawscore21$pitcherthrows=="R" & datawscore21$pitcherthrows=="L")

sum(datawscore21$batterside=="R")
sum(datawscore21$batterside=="L")
sum(datawscore21$batterside=="R" & datawscore21$batterside=="L")

#finding the frequency of each pitch type and their result
datawscore21_count <- datawscore21 %>%
  group_by(pitchname, pitchresult, pitcherthrows, batterside) %>%
  summarize(count = n())  # Calculate the frequency of each combination

ggplot(datawscore21_count, aes(x = pitchname, y = pitchresult, size = count)) +
  geom_point(alpha = 0.6) +  
  scale_size_continuous(range = c(1, 10))

ggplot(datawscore21_count, aes(x=pitchname, y= pitchresult, size =count)) + geom_point(alpha = 0.3)+
    scale_size(range = c(.1, 24)) #overall

ggplot(datawscore21_count, aes(x=pitchname, y= pitchresult, size =count)) +
  geom_point(alpha = 0.3)+
  scale_size(range = c(.1, 24))+ 
  facet_grid(pitcherthrows~batterside) #impact of handedness
```

#### Closer B/S Count Pitch Types

```{r}
total_counts <- all_closer_data %>%
  group_by(balls, strikes) %>%
  summarise(total = n(), .groups = "drop")

# Aggregate the data by pitch type, ball, and strike count
aggregated_data <- all_closer_data %>%
  group_by(pitchname_desc, balls, strikes) %>%
  summarise(count = n(), .groups = "drop") %>%
  left_join(total_counts, by = c("balls", "strikes")) %>%  # Join to get total counts
  mutate(proportion = count / total)  # Calculate proportion

pitch_types <- unique(aggregated_data$pitchname_desc)

for (pitch_type in pitch_types) {
  # Create the plot for the current pitch type
  plot <- ggplot(aggregated_data %>% filter(pitchname_desc == pitch_type), 
                 aes(x = balls, y = strikes, fill = proportion)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "red") +
    labs(
      title = paste0("Heatmap of ", pitch_type, " by Ball and Strike Count (Proportions)"),
      x = "Ball Count",
      y = "Strike Count",
      fill = "Proportion"
    ) +
    theme_minimal()

  # Display the plot (optional)
  print(plot)
}
```

## Reliever EDA

#### Creating Dataset

```{r}
# Group each individual game together (by game id or game date)
# Get all pitchers for home and away team in order
# Remove first pitcher and then ensure none of datapoints in closer set.
# Minimum threshold for number of appearances to exclude teams using starting pitchers.
# 2021 Relievers
pitch_order_2021 <- pitchwithscore21 %>%
  group_by(gameid, teambat) %>%
  arrange(ab) %>%  # Arrange by the order of at-bats or appearances
  summarise(
    pitcher_order = list(unique(pitcher)[-1])
  )
# Filtering the original dataset based on the pitcher list
reliever_2021 <- pitchwithscore21 %>%
  group_by(gameid, teambat) %>%
  filter(pitcher %in% unlist(pitch_order_2021$pitcher_order[match(gameid, pitch_order_2021$gameid)])) %>% # make sure it's a pitcher in this list. 
   filter(!(pitcher %in% espnbest21)) %>% ungroup()
# 2022 Relievers
pitch_order_2022 <- pitchwithscore22 %>%
  group_by(gameid, teambat) %>%
  arrange(ab) %>%  # Arrange by the order of at-bats or appearances
  summarise(
    pitcher_order = list(unique(pitcher)[-1])
  )
# Filtering the original dataset based on the pitcher list
reliever_2022 <- pitchwithscore22 %>%
  group_by(gameid, teambat) %>%
  filter(pitcher %in% unlist(pitch_order_2022$pitcher_order[match(gameid, pitch_order_2022$gameid)])) %>% # make sure it's a pitcher in this list. 
   filter(!(pitcher %in% espnbest22)) %>% ungroup()
# 2023 Relievers
pitch_order_2023 <- pitchwithscore23 %>%
  group_by(gameid, teambat) %>%
  arrange(ab) %>%  # Arrange by the order of at-bats or appearances
  summarise(
    pitcher_order = list(unique(pitcher)[-1])
  )
# Filtering the original dataset based on the pitcher list
reliever_2023 <- pitchwithscore23 %>%
  group_by(gameid, teambat) %>%
  filter(pitcher %in% unlist(pitch_order_2023$pitcher_order[match(gameid, pitch_order_2023$gameid)])) %>% # make sure it's not a pitcher in this list. 
   filter(!(pitcher %in% espnbest23)) %>% ungroup()

# Combined Data
all_reliever_data<- rbind(reliever_2021, reliever_2022, reliever_2023) %>% select(-X)
  
```

#### Reliever Pitch Types

```{r}
ggplot(data=all_reliever_data, mapping = aes(x = pitchname, fill = pitchname))  +
  geom_bar() + xlab("Pitch Type") + ylab("Frequency") + ggtitle("Pitch Type Frequency for Relievers from 2021-2023")
out_reliever_pitch <- all_reliever_data %>% filter(pitchresult %in% strike_out)
ggplot(data=out_reliever_pitch, mapping = aes(x = pitchname, fill = pitchname))  + geom_bar() + xlab("Pitch Type") + ylab("Frequency") + ggtitle("Relievers Pitch Type Frequency Resulting in Strikeout/Outs")
# Four-seam FB
all_reliever_ff <- filter(all_reliever_data, pitchname=="FF")
boxplot(scale(select(all_reliever_ff, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Fastball Pitching Stats Boxplots")
# Sinker
all_reliever_si <- filter(all_reliever_data, pitchname=="SI")
boxplot(scale(select(all_reliever_si, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Sinker Pitching Stats Boxplots")
# Slider
all_reliever_sl <- filter(all_reliever_data, pitchname=="SL")
boxplot(scale(select(all_reliever_sl, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Slider Pitching Stats Boxplots")
# Curveball
all_reliever_cu <- filter(all_reliever_data, pitchname=="CU")
boxplot(scale(select(all_reliever_cu, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Curveball Pitching Stats Boxplots")
# Changeup
all_reliever_ch <- filter(all_reliever_data, pitchname=="CH")
boxplot(scale(select(all_reliever_ch, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Changeup Pitching Stats Boxplots")
# Cutter
all_reliever_fc <- filter(all_reliever_data, pitchname=="FC")
boxplot(scale(select(all_reliever_fc, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Cutter Pitching Stats Boxplots")
# Splitter
all_reliever_fs <- filter(all_reliever_data, pitchname=="FS")
boxplot(scale(select(all_reliever_fs, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Splitter Pitching Stats Boxplots")
# Knuckle Curve
all_reliever_kc <- filter(all_reliever_data, pitchname=="KC")
boxplot(scale(select(all_reliever_kc, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Knuckle Curve Pitching Stats Boxplots")
# Sweeper
all_reliever_st <- filter(all_reliever_data, pitchname=="ST")
boxplot(scale(select(all_reliever_st, spinrate, relspeed, horzbreak, inducedvertbreak, platelocside, platelocheight)), ylab = "Standardized Values", las = 2, cex.axis = 0.8)
title("Sweeper Pitching Stats Boxplots")
```

```         
```

#### Reliever B/S Count

```{r}
total_counts <- all_reliever_data %>%
  group_by(balls, strikes) %>%
  summarise(total = n(), .groups = "drop")

# Aggregate the data by pitch type, ball, and strike count
aggregated_data <- all_reliever_data %>%
  group_by(pitchname_desc, balls, strikes) %>%
  summarise(count = n(), .groups = "drop") %>%
  left_join(total_counts, by = c("balls", "strikes")) %>%  # Join to get total counts
  mutate(proportion = count / total)  # Calculate proportion

pitch_types <- unique(aggregated_data$pitchname_desc)

for (pitch_type in pitch_types) {
  # Create the plot for the current pitch type
  plot <- ggplot(aggregated_data %>% filter(pitchname_desc == pitch_type), 
                 aes(x = balls, y = strikes, fill = proportion)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "red") +
    labs(
      title = paste0("Heatmap of ", pitch_type, " by Ball and Strike Count (Proportions)"),
      x = "Ball Count",
      y = "Strike Count",
      fill = "Proportion"
    ) +
    theme_minimal()

  # Display the plot (optional)
  print(plot)
}
```

#### Reliever Individual Statistics

```{r}
# Way 1: Reliever Pitch Type Averages per row
pitcherName <- paste0(all_reliever_data$pitcher, " table")
reliever_tables <- split(all_reliever_data, pitcherName)

# Create an empty data frame to store averages
median_by_reliever <- data.frame(name=character(), pitch=character(), median_spin=numeric(), median_velo=numeric(),
                              median_hbreak=numeric(), median_vbreak=numeric(), median_plside=numeric(),
                              median_plheight=numeric(), stringsAsFactors=FALSE)

# Get Average Stats for Each Pitcher's Pitch Types

# Loop through each reliever
for (i in reliever_tables) {
  pitchtypes <- unique(i$pitchname)

  # Split by pitch type
  table_by_pitch <- split(i, i$pitchname)

  # Loop through each pitch type and calculate averages
  for (j in table_by_pitch) {
    median_spin <- median(j$spinrate, na.rm=TRUE)
    median_velo <- median(j$relspeed, na.rm=TRUE)
    median_hbreak <- median(j$horzbreak, na.rm=TRUE)
    median_vbreak <- median(j$inducedvertbreak, na.rm=TRUE)
    median_plside <- median(j$platelocside, na.rm=TRUE)
    median_plheight <- median(j$platelocheight, na.rm=TRUE)

    # Append the data to median_by_reliever
    median_by_reliever <- rbind(median_by_reliever,
                             data.frame(name=i$pitcher[1], pitch=j$pitchname[1], median_spin=median_spin,
                                        median_velo=median_velo, median_hbreak=median_hbreak, median_vbreak=median_vbreak,
                                        median_plside=median_plside, median_plheight=median_plheight))
  }
}

# plot
ggplot(median_by_reliever, aes(x=pitch, y=median_vbreak)) +
  geom_boxplot() +
  ggtitle("Reliever Boxplots of Vertical Break by Pitch Type")

```

#### Identify Number of Relievers

```{r}
relievers <- unique(median_by_reliever$name)
num_relievers <- length(relievers)
```

#### One-Way ANOVA comparing statistics by division

```{r}
# Creating a data frame with pitcher names and their respective teams for 2021, 2022, and 2023
pitcher_data <- data.frame(
  pitcher = c("Melancon, Mark", "Hendriks, Liam", "Jansen, Kenley", "Smith, Will", 
              "Iglesias, Raisel", "Hader, Josh", "Diaz, Edwin", "McGee, Jake", 
              "Chapman, Aroldis", "Reyes, Alex", "Pressly, Ryan", "Clase, Emmanuel", 
              "Barnes, Matt", "Romano, Jordan", "Kimbrel, Craig", "Trivino, Lou", 
              "Hand, Brad", "Bard, Daniel", "Soto, Gregory", "Colome, Alex", 
              "Barlow, Scott", "Kennedy, Ian", "Floro, Dylan", "Steckenrider, Drew", 
              "Rodriguez, Richard", "Castillo, Diego", "Neris, Hector", "Sulser,                    Cole", "Hembree, Heath", "Soria, Joakim", 
              # 2022 data
              "Clase, Emmanuel", "Jansen, Kenley", "Hendriks, Liam", "Romano, Jordan",
              "Bard, Daniel", "Pressly, Ryan", "Diaz, Edwin", "Soto, Gregory", 
              "Hader, Josh", "Rogers, Taylor", "Doval, Camilo", "Barlow, Scott", 
              "Kimbrel, Craig", "Sewald, Paul", "Holmes, Clay", "Scott, Tanner", 
              "Helsley, Ryan", "Bednar, David", "Lopez, Jorge", "Melancon, Mark", 
              "Iglesias, Raisel", "Robertson, David", "Barlow, Joe", "Knebel, Corey", 
              "Rainey, Tanner", "Jimenez, Dany", "Diaz, Alexis", "Pagan, Emilio", 
              "Schreiber, John", "Adam, Jason",
              # 2023 data
              "Clase, Emmanuel", "Bednar, David", "Doval, Camilo", "Diaz, Alexis", 
              "Romano, Jordan", "Williams, Devin", "Bautista, Felix", "Hader, Josh", 
              "Iglesias, Raisel", "Pressly, Ryan", "Estevez, Carlos", "Jansen, Kenley",
              "Finnegan, Kyle", "Duran, Jhoan", "Lange, Alex", "Fairbanks, Pete", 
              "Holmes, Clay", "Phillips, Evan", "Kimbrel, Craig", "Alzolay, Adbert", 
              "Smith, Will", "May, Trevor", "Sewald, Paul", "Puk, A.J.", 
              "Robertson, David", "Helsley, Ryan", "Johnson, Pierce", "Barlow, Scott", 
              "Sewald, Paul", "Graveman, Kendall"),
  team = c("Padres", "White Sox", "Dodgers", "Braves", 
           "Angels", "Brewers", "Mets", "Giants", 
           "Yankees", "Cardinals", "Astros", "Guardians", 
           "Red Sox", "Blue Jays", "Cubs", "Athletics", 
           "Nationals", "Rockies", "Tigers", "Twins", 
           "Royals", "Rangers", "Marlins", "Mariners", 
           "Pirates", "Rays", "Phillies", "Orioles", 
           "Reds", "Diamondbacks",
           # 2022 teams
           "Guardians", "Braves", "White Sox", "Blue Jays", 
           "Rockies", "Astros", "Mets", "Tigers", 
           "Brewers", "Padres", "Giants", "Royals", 
           "Dodgers", "Mariners", "Yankees", "Marlins", 
           "Cardinals", "Pirates", "Orioles", "Diamondbacks", 
           "Angels", "Cubs", "Rangers", "Phillies", 
           "Nationals", "Athletics", "Reds", "Twins", 
           "Red Sox", "Rays", 
           # 2023 teams
           "Guardians", "Pirates", "Giants", "Reds", 
           "Blue Jays", "Brewers", "Orioles", "Padres", 
           "Braves", "Astros", "Angels", "Red Sox", 
           "Nationals", "Twins", "Tigers", "Rays", 
           "Yankees", "Dodgers", "Phillies", "Cubs", 
           "Rangers", "Athletics", "Mariners", "Marlins", 
           "Mets", "Cardinals", "Rockies", "Royals", 
           "Diamondbacks", "White Sox")
)

# Assigning divisions based on teams
pitcher_data$division[pitcher_data$team %in% c("Yankees", "Red Sox", "Blue Jays", "Rays", "Orioles")] <- "AL East"
pitcher_data$division[pitcher_data$team %in% c("White Sox", "Guardians", "Tigers", "Royals", "Twins")] <- "AL Central"
pitcher_data$division[pitcher_data$team %in% c("Astros", "Angels", "Athletics", "Mariners", "Rangers")] <- "AL West"
pitcher_data$division[pitcher_data$team %in% c("Braves", "Marlins", "Mets", "Phillies", "Nationals")] <- "NL East"
pitcher_data$division[pitcher_data$team %in% c("Cubs", "Reds", "Brewers", "Pirates", "Cardinals")] <- "NL Central"
pitcher_data$division[pitcher_data$team %in% c("Dodgers", "Giants", "Padres", "Diamondbacks", "Rockies")] <- "NL West"

# Function to assign team and division based on pitcher names
assign_team_division <- function(closers_data, mapping_data) {
  # Merge the closers data with the mapping data based on the pitcher names
  merged_data <- merge(closers_data, mapping_data, by = "pitcher", all.x = TRUE)
  
  # Select the relevant columns, including the division from the mapping data
  merged_data$division <- merged_data$division.y  # Use division from mapping data
  merged_data <- merged_data[, c(names(closers_data), "division")]  # Keep only relevant columns
  
  return(merged_data)
}

# Merging teams and divisions directly
espnsbestclosers21 <- merge(espnsbestclosers21, mapping_data, by = "pitcher", all.x = TRUE)
espnsbestclosers22 <- merge(espnsbestclosers22, mapping_data, by = "pitcher", all.x = TRUE)
espnsbestclosers23 <- merge(espnsbestclosers23, mapping_data, by = "pitcher", all.x = TRUE)

# If the division is named division.y after the merge, rename it
colnames(espnsbestclosers21)[which(colnames(espnsbestclosers21) == "division.y")] <- "division"
colnames(espnsbestclosers22)[which(colnames(espnsbestclosers22) == "division.y")] <- "division"
colnames(espnsbestclosers23)[which(colnames(espnsbestclosers23) == "division.y")] <- "division"


# Initialize a list to store ANOVA results
anova_results <- list()
variables_to_analyze <- c("spinrate", "relspeed", "horzbreak", "inducedvertbreak", "platelocside", "platelocheight")

# Perform ANOVA for each variable across divisions for each year 
for (var in variables_to_analyze) {
  year_results <- list()
  
  for (year_data in list(espnsbestclosers21, espnsbestclosers22, espnsbestclosers23)) {
    anova_model <- aov(as.formula(paste(var, "~ division")), data = year_data)
    year_results[[length(year_results) + 1]] <- summary(anova_model)
  }
  
  # Store results for the variable
  anova_results[[var]] <- year_results
}

# View the results
anova_results

```

For all metrics, the p-values are extremely low (all < 2e-16), indicating that there is a statistically significant difference in the means of these metrics across the divisions for each year. This suggests that the division a pitcher belongs to affects these metrics significantly.
The results suggest that divisions significantly impact these pitching metrics. However, the degree of impact appears to fluctuate over the years. This analysis could guide further investigations, such as post-hoc tests (e.g., Tukey's HSD) to identify which specific divisions differ from each other within each metric.

#### ANOVA comparing closer and reliever metrics

```{r}
# Define the datasets and variables to analyze
datasets <- list(
  closers21 = espnsbestclosers21,
  closers22 = espnsbestclosers22,
  closers23 = espnsbestclosers23,
  relievers21 = reliever_2021,
  relievers22 = reliever_2022,
  relievers23 = reliever_2023
)

variables_to_analyze <- c("spinrate", "relspeed", "horzbreak", "inducedvertbreak", "platelocside", "platelocheight")

# Initialize a list to store ANOVA results
anova_results <- list()

# Loop through each variable and perform ANOVA for each year
for (var in variables_to_analyze) {
  # Create a data frame to hold data for ANOVA
  combined_data <- data.frame(
    value = numeric(0),
    group = character(0)
  )
  
  # Extract data for each year
  for (year in c("21", "22", "23")) {
    # Get closers data
    closers_data <- datasets[[paste0("closers", year)]][, var, drop = FALSE]
    closers_data$group <- "Closer"
    
    # Get relievers data
    relievers_data <- datasets[[paste0("relievers", year)]][, var, drop = FALSE]
    relievers_data$group <- "Reliever"
    
    # Combine data
    combined_data <- rbind(combined_data, 
                            data.frame(value = closers_data[[var]], group = closers_data$group),
                            data.frame(value = relievers_data[[var]], group = relievers_data$group))
  }
  
  # Perform ANOVA
  anova_results[[var]] <- aov(value ~ group, data = combined_data)
}

# Print ANOVA summary for each variable
for (var in variables_to_analyze) {
  cat(paste("ANOVA results for", var, ":\n"))
  print(summary(anova_results[[var]]))
  cat("\n")
}

```

All variables show statistically significant differences (p < 0.05) between the groups, indicating that the pitching characteristics of closers and relievers differ.The differences could be attributed to the different roles and strategies employed by closers and relievers in games. Closers are often tasked with finishing games, which may lead them to use different pitch strategies or techniques compared to relievers, who may have varying roles and responsibilities throughout the game. 
